Subsidiary Recursion in Coq

### 2022-01-22 paper/paper.tex
Tiny, Tiny code fix in paper.
revising paper based on Chris's feedback (more to come)

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We should emphasize that
this is a derivation within the type theory of Coq, and requires no
axioms or other modifications to Coq. Using this derivation, we
present several example functions like \verb|wordsBy|, and prove
theorems about them.  A nice example is a definition of run-length
encoding using \verb|span| as a subsidiary recursion, where we prove
that encoding and then decoding returns the original list.  Our
approach applies to the standard datatypes in the Coq library, and
does not require switching libraries or datatype definitions.

An important technical novelty is a derivation of a weakened form of
positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
datatypes $D$ to be strictly positive: in the type for any constructor
of $D$, $D$ cannot occur to the left of any arrows.  Our derivation
needs to use positive-recursive types, where $D$ may occur to the left
of an even number (only) of arrows.  We present a way to derive a
weakened form of positive-recursive type that is sufficient for our
examples (Section~\ref{sec:mu}).  The weakening is to require only
that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
and $F\ \mu$ its one-step unfolding.  Usually these types are
isomorphic.  Hence, we dub these \textbf{retractive-positive}
recursive types.  This weakening leads to noncanonical elements of
$\mu$, but we will see how to work around this.  Our definition of
retractive-positive recursive types makes essential use of impredicative
quantification, and hence is not legal in predicative theories
like Agda's.

@@

All presented derivations have been checked with Coq version 8.13.2,
using command-line option \verb|-impredicative-set|.  The code may be
found as release \verb|itp-2022| (dated prior to the ITP 2022
deadline) at \url{https://github.com/astump/coq-subsidiary}.  The
paper references files in this codebase, as an aid to the reader
wishing to peruse the code.

@@

Algebras for our subsidiary recursion are more complex than $F$-algebras.  First, for
reasons we will explain further below, the carrier of the algebra will
be a functor \verb|X : Set -> Set|.  Second, algebras have a specified
\emph{anchor type} \verb|C|, which we can think of as the datatype
\emph{as viewed by a containing recursion} or else, if this is a
top-level recursion, our development's version of the actual datatype
(e.g., \verb|List|).  The algebra is presented with:

@@

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have an entity (here, \verb|R|) from which one cannot
escape by using the available operations (for other cases:
the ordinal $\epsilon_0$ and $\omega^-$, the physical universe and traveling at
the speed of light).  Staying in the recursion universe is good,
because we may recurse (via \verb|eval|) on any value of type
\verb|R|.  Some points must still be explained: why \verb|X| has type \verb|Set -> Set|,
and the definition of \verb|FoldT|.  Let us see these details next.

@@

\verb|FoldT Alg C| is the type for fold functions which apply
algebras of type \verb|Alg| to data of type \verb|C|, which we have
already dubbed the \emph{anchor type} of the recursion.  At the top
level of code, the anchor type would just be \verb|List| (for
example).  When one initiates a subsidiary recursion, though, the
anchor type will instead by the abstract type \verb|R| for the outer recursion.
The variable \verb|Alg| occurs only positively (but not strictly
positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg C X|.  An algebra can assume nothing about \verb|R|
except that it supports the following operations.  First there is
\verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
that the data of type \verb|R| are really values of the anchor type
of this recursion.  Next we have \texttt{fold}, which will allow us to fold
another algebra over data of type \verb|R|.  We will use \verb|fold|
to initiate subsidiary recursions.  Then there is \verb|eval|, for
recursive calls on data of type \verb|R|.

@@

Finally, Figure~\ref{fig:subrecb} defines \verb|out| as a subsidiary
recursion, given any fold function with its anchor type \verb|R|.  The
code for \verb|out| just folds an algebra over the input of type
\verb|R|, where that algebra simply returns the subdata structure it
is given.  Outside the recusion, this has type \verb|F R|; inside the
recursion it has type \verb|F R'| where \verb|R'| is the abstract type
of the subsidiary recursion (named in the figure just for discussion
here).  So \verb|out| implements the idea that unfolding an abstract
type one step is just a trivial case of subsidiary recursion.

@@

Our work contributes to the program proposed by Owens and Slind, of
broadening the scope of functional programs that can be accommodated
in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
advocated in the literature on programming languages under the name
\emph{strong functional programming}~\cite{turner95}.  Uustalu and
Vene developed a categorical view of a recursion scheme allowing one
level of subsidiary recursion, and illustrated it with an artificial
example~\cite{uustalu11}.  In contrast, our scheme allows arbitrary
finite nestings of recursion, and we illustrate it with realistic
examples drawn from Haskell's \verb|Data.List.Extra|.

\textbf{Mendler-style recursion.}  Mendler introduced the idea of using
universal abstraction to support compositional termination checking~\cite{mendler91}.  He proposed a
functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
    \to X}) \to \mu\ F\ \to X}$. We have adopted this idea to the
constructor of the type \verb|Mu| (Section~\ref{sec:mu}).  Previous
works explored the categorical perspective on Mendler-style
recursion~\cite{uustalu99}, and its use with negative type
schemes~\cite{ahn11}.  Previous work from our group showed
how to derive inductive datatypes in Cedille using encodings
extending the Mendler encoding~\cite{firsov+18b,firsov+18a}.
Here, we do not derive inductive types using these methods,
but rather apply them to justify a terminating recursion scheme.

>>>>

This paper presents a functor-generic derivation of terminating
subsidiary recursion and induction in Coq.  We emphasize that this is
a derivation within the type theory of Coq, and requires no axioms or
other modifications to Coq, except the \verb|-impredicative-set|
flag. Using this derivation, we present several example functions like
\verb|wordsBy|, and prove theorems about them.  A nice example is a
definition of run-length encoding using \verb|span| as a subsidiary
recursion, where we prove that encoding and then decoding returns the
original list.  Our approach applies to the standard datatypes in the
Coq library, and does not require switching libraries or datatype
definitions.

An important technical novelty is a derivation of a weakened form of
positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
datatypes $D$ to be strictly positive: in the input types of
constructors of $D$, $D$ cannot occur to the left of any arrows.  Our
derivation needs to use positive-recursive types, where $D$ may occur
to the left of an even number (only) of arrows.  We present a way to
derive a weakened form of positive-recursive type that is sufficient
for our examples (Section~\ref{sec:mu}).  The weakening is to require
only that $F\ (\mu F)$ is a retract of $\mu F$.  Usually these types are
isomorphic.  Hence, we dub these \textbf{retractive-positive}
recursive types.  This weakening leads to noncanonical elements of
$\mu$, but we will see how to work around this.  Our definition of
retractive-positive recursive types makes essential use of
impredicative quantification, and hence is not legal in predicative
theories like Agda's.

@@

All presented derivations have been checked with Coq version 8.13.2.  The code may be
found as release \verb|itp-2022| (dated prior to the ITP 2022
deadline) at \url{https://github.com/astump/coq-subsidiary}.  The
paper references files in this codebase, as an aid to the reader
wishing to peruse the code.

@@

Algebras for our subsidiary recursion are more complex than $F$-algebras.  First, for
reasons we will explain further below, the carrier of the algebra will
be a functor \verb|X : Set -> Set|.  Second, algebras have a specified
\emph{anchor type} \verb|C|, which we can think of as the datatype
\emph{as viewed by a containing recursion} or else, if this is a
top-level recursion, our development's version of the actual datatype
(e.g., \verb|List A|, for some \verb|A|).  The algebra is presented with:

@@

As in previous work, we dub this interface a \emph{recursion
universe}~\cite{stump20}.  As in other domains using the term
``universe'', we have a kind of space (here, \verb|R|) with operations
that keep one in that space (for other cases: the ordinal $\epsilon_0$
and $\omega^-$, the physical universe and traveling at the speed of
light).  Staying in the recursion universe is good, because we may
recurse (via \verb|eval|) on any value of type \verb|R|.  One can
use \verb|reveal| to leave, but then \verb|eval| can no longer be used. Some points
must still be explained: why \verb|X| has type \verb|Set -> Set|, and
the definition of \verb|FoldT|.  Let us see these details next.

@@

\verb|FoldT Alg C| is the type for fold functions which apply
algebras (\verb|Alg|) to data of type \verb|C|, which we have already
dubbed the \emph{anchor type} of the recursion.  At the top level of
code, the anchor type would just be \verb|List| (for example).  When
one initiates a subsidiary recursion, though, the anchor type will
instead by the abstract type \verb|R| for the outer recursion.  The
variable \verb|Alg| occurs only positively (but not strictly
positively) in \verb|AlgF|, because it occurs negatively in
\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
we can indeed take a fixed-point of \verb|AlgF| to define the constant
\verb|Alg|.

Let us look at \verb|AlgF|.  As noted already, each recursion is based
on an abstract type \verb|R|, representing the data upon which we will
recurse.  This is the first argument to a value of type
\verb|AlgF Alg C X|.  Reasoning parametrically, an algebra can assume
nothing about \verb|R| except that it supports the following
operations.  First there is \verb|reveal|, which turns an \verb|R|
into a \verb|C|.  This reveals that the data of type \verb|R| are
really values of the anchor type of this recursion.  Next we have
\texttt{fold}, which will allow us to fold another algebra over data
of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
recursions.  Then there is \verb|eval|, for recursive calls on data of
type \verb|R|.

@@

Finally, Figure~\ref{fig:subrecb} defines \verb|out| as a subsidiary
recursion, given any fold function with its anchor type \verb|R|.
Outside the recursion, \verb|d| has type \verb|F R|; inside the
recursion it has type \verb|F R'| where \verb|R'| is the abstract type
of the subsidiary recursion.  So \verb|out| implements the idea that
unfolding an abstract type one step is just a trivial case of
subsidiary recursion.

@@

Our work contributes to the program proposed by Owens and Slind, of
broadening the scope of functional programs that can be accommodated
in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
advocated in the literature on programming languages under the name
\emph{strong functional programming}~\cite{turner95}.  Uustalu and
Vene developed a categorical view of a recursion scheme allowing one
level of subsidiary recursion, and illustrated it in Haskell with an
artificial example~\cite{uustalu11}.  In contrast, our scheme allows
arbitrary finite nestings of recursion, and we illustrate it in Coq
with realistic examples.  Like them, we find that subsidiary recursion
subsumes course-of-value recursion.

\textbf{Mendler-style recursion.}  Mendler introduced the idea of using
universal abstraction to support compositional termination checking~\cite{mendler91}.  He proposed a
functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
    \to X}) \to \mu\ F\ \to X}$. We have adopted this idea to the
constructor of the type \verb|Mu| (Section~\ref{sec:mu}).  Previous
works explored the categorical perspective on Mendler-style
recursion~\cite{uustalu99}, and its use with negative type
schemes~\cite{ahn11}.  Previous work from our group showed
how to derive inductive datatypes in Cedille using encodings
extending the Mendler encoding~\cite{firsov+18b,firsov+18a}.
Here, we do not derive inductive types using these methods,
but rather apply them to justify a terminating recursion scheme
for existing datatypes.
